surfing

### 2021-05-07 inst/surfing.Rmd
citations

---
title: "surfing"
output: bookdown::html_document2
---

@@

@@

Surfing is a popular and growing activity in terms of both
participation and viewing.  Competitive surfing involves surfers
competing against one or two other surfers in heats lasting 30-50
minutes.  Points are awarded on a 10-point scale: each of five judges
awards points to a surfer riding a particular wave.

The typical format is as follows.  Up to four surfers---the
competitors---are in the water simultaneously, watched by up to five
judges.  Subject to surfing etiquette, competitors are free to catch a
wave at their discresion within the heat's time window, typically
20-30 minutes.  The judges rate each wave ride and award points to the
surfer based on a range of subjective criteria including commitment
and degree of difficulty, innovation and progress of manoeuvres,
variety of manoeuvres and speed, power and flow.  The top score and
bottom score among the five judges are removed and the remaining three
judges' scores are averaged to give the final score for the surfer for
that wave.  Each surfer's aggregate score is given by the average of
the two highest-scoring waves.  The winner of the tournament is the
surfer with the highest aggregate score.

@@

The World Surf League (WSL) is the main governing body of surfing
competitions.  The WSL conducts a world tour in which the best ranked
34 surfers compete; in addition, at each tour venue, two "wildcard"
surfers also enter the competition.  The 2019 WSL tour saw competitive
events held at eleven different surf locations.  In each of these
locations the wave type of wave differs.  Some events are known to be
more favourable to those who have a high number of manoeuvres at their
disposal and others are more suited to surfers who regularly surf
bigger waves.

>>>>

---
title: "surfing"
output: bookdown::html_document2
bibliography: chess.bib
---

@@

@fischer1969

[@fischer1969]

[-@fischer1969]

@@

Surfing is a popular and growing activity in terms of both
participation and viewing.  Competitive surfing involves surfers
competing against one or two other surfers in heats lasting 30-50
minutes.  Points are awarded on a 10-point scale: each of five judges
awards points to a surfer riding a particular wave.  The World Surf
League (WSL) is the main governing body of surfing competitions.  The
WSL conducts a world tour in which the best ranked 34 surfers compete;
in addition, at each tour venue, two "wildcard" surfers also enter the
competition, who are ignored here.  The 2019 WSL tour saw competitive
events held at eleven different surf locations.

The typical format is as follows.  Up to four surfers---the
competitors---are in the water simultaneously, watched by up to five
judges.  Subject to surfing etiquette, competitors are free to catch a
wave at their discretion within the heat's time window, typically
20-30 minutes.  The judges rate each wave ride and award points to the
surfer based on a range of subjective criteria including commitment
and degree of difficulty, innovation and variety of manoeuvres, and
transition between manoeuvres.  The top score and bottom score among
the five judges are removed and the remaining three judges' scores are
averaged to give the final score for the surfer for that wave.  Each
surfer's aggregate score is given by the average of the two
highest-scoring waves.  The winner of the tournament is the surfer
with the highest aggregate score.

@@

# Bradley-Terry

\subsection{Bradley-Terry and generalizations for rank
statistics}\label{bradley-terry-and-generalizations-for-rank-statistics}

The Bradley-Terry model [@bradley1954] assigns non-negative strengths
$p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
probability of $i$ beating $j\neq i$ in pairwise competition is
$\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
p_i=1$.  Further, we use a generalization due to [@luce1959], in which
the probability of competitor~$i$ winning in a field of $\left\lbrace
1,\ldots, n\right\rbrace$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
that there is information in the whole of the finishing order, and not
just the first across the line, we can follow [@plackett1975] and
consider the runner-up to be the winner among the remaining
competitors, and so on down the finishing order. Without loss of
generality, if the order of finishing were $1,2,3,4,5$, then a
suitable Plackett-Luce likelihood function would be

\begin{equation}\label{competitors_1_to_5_likelihood}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}\cdot
\frac{p_4}{p_4+p_5}\cdot
\frac{p_5}{p_5}
\end{equation}

and this would be a forward ranking Plackett-Luce model in the
terminology of [@johnson2020].  A slight generalization allows the
incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
did not finish, we would have

\begin{equation}\label{competitors_1_to_3_only_finished}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}
\end{equation}

(observe how this likelihood function, while informative about
$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$).  We now
use a technique due to Hankin (2010, 2017) and introduce
fictional (reified) entities whose nonzero Bradley-Terry strength
helps certain competitors or sets of competitors under certain
conditions. The canonical example would be the home-ground advantage
in association football.  If players (teams) $1,2$ with strengths
$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
away wins for team $1$, and $c$ home wins and $d$ away wins for
team~$2$, then a suitable likelihood function would be

\[
\left(\frac{p_1+p_H}{p_1+p_2+p_H}\right)^a
\left(\frac{p_1}{p_1+p_2+p_H}\right)^b
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^c
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^d,
\]

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and effective coordination between
members of doubles tennis teams; we may use a similar device to
account for (e.g.) wet conditions in Formula 1.

# Literature Review and previous work

Most previous research into competitive surfing is limited to the
scoring of ariel manoeuvres compared with other manoeuvres in events
[10] and analysis in the difference in average wave scores for the top
10 and bottom 10 surfers using a One-way ANOVA method [11].
